---
title: "Chapter 3 -  Data Manipulation"
output: html_document
---

## Data Already in R

Run once to import libraries:
```{r}
library("ggplot2")
library("magrittr")
```

The __Dataset__ package:

```{r}
library(datasets)
library(help = "datasets")

data(cars)
head(cars)
cars %>% qplot(speed, dist, data = .)
```

Another package with useful datasets is *mlbench*, it can be installed:

```{r}
# install.packages("mlbench")
library(mlbench)
library(help = "mlbench")
```

## Quickly Reviewing Data

Function `head()`, shows the first $n$ lines.

```{r}
cars %>% head(3)
```

Function `tail()`, shows the last $n$ lines.

```{r}
cars %>% tail(3)
```

To get summary statistics for all the columns in a data frame, use the 
`summary()` function:

```{r}
cars %>% summary
```

The Iris dataset:

```{r}
data(iris)
iris %>% summary
```

## Reading data

Reading data in table:

```{r}
?read.table
```

## Examples of Reading and Formatting datasets

```{r}
library(mlbench)
data(BreastCancer)
BreastCancer %>% head(3)
```

## Manipulating Data with `dplyr`

### Some Sueful `dplyr` Functions

```{r}
# install.packages("dplyr")
library("dplyr")
iris %>% as_tibble()
```

#### select(): Pick Selected Columns and Get Rid of The Rest

```{r}
iris %>% as_tibble() %>% select(Petal.Width) %>% head(3)
iris %>% as_tibble() %>% select(Sepal.Length:Petal.Length)
```

Pick columns based on names:

```{r}
iris %>% as_tibble() %>% select(starts_with("Petal")) %>% head(3)
iris %>% as_tibble() %>% select(ends_with("Width")) %>% head(3)
iris %>% as_tibble() %>% select(contains("etal")) %>% head(3)
iris %>% as_tibble() %>% select(matches(".t.")) %>% head(3)
```

#### mutate(): Add computed values to your Data Frame 

The `mutate()` function lets you add a column to your data frame by specifying an expression
for how to compute it:

```{r}
iris %>% as_tibble() %>% 
    mutate(Petal.Width.plus.Length = Petal.Width + Petal.Length) %>%
    select(Species, Petal.Width.plus.Length) %>%
    head(3)
```

You can add more columns than one by specifying them in the `mutate()` function:

```{r}
iris %>% as_tibble() %>% 
    mutate(Petal.Width.plus.Length = Petal.Width + Petal.Length,
           Sepal.Width.plus.Length = Sepal.Width + Sepal.Length) %>%
    select(Petal.Width.plus.Length, Sepal.Width.plus.Length) %>%
    head(3)
```

#### Transmute: Add computed values to your data frame and get rid of all other columns

The `transmute()` function works just like the `mutate()` function, except 
is combined with a select() so the results is a data frame that on;y contains the new columns you make.

```{r}
iris %>% as_tibble() %>%
    transmute(Petal.Width.plus.Length = Petal.Width + Petal.Length) %>%
    head(3)
```

#### arrange(): Reorder your data frame by Sorting Columns

The `arrange()` function just reorders the data frame by sorting columns acoording to 
what specify:

```{r}
iris %>% as_tibble() %>%
    arrange(Sepal.Length) %>%
    head(3)
```

Decreasing order:

```{r}
iris %>% as_tibble() %>%
    arrange(desc(Sepal.Length)) %>%
    head(3)
```

#### filter(): Pick selected Rows and Get Rid of the Rest

The `filter()` functions lets you pick out rows based on logical expressions.
You give the functions a predicate, specifying what a row should satisfy to be included.

```{r}
iris %>% as_tibble() %>%
    filter(Sepal.Length > 5) %>%
    head(3)
```

More complex logical expressions are accepted:


```{r}
iris %>% as_tibble() %>%
    filter(Sepal.Length > 5 & Species == "virginica") %>%
    head(3)
```

#### group_by(): Split tour Data Into Subtables Based on Column Values 
The `group_by()` function tells `dplyr` that you want to work on dat separated into different subsets.

```{r}
iris %>% as_tibble() %>% group_by(Species) %>% head(3)
```

#### summarise/summarize(): Calculate Summary Statistics

```{r}
iris %>%
    summarise(Mean.Petal.Length = mean(Petal.Length),
              Mean.Sepal.Length = mean(Sepal.Length))
```

Where this is really powerful is in combination with `group_by()`. There you can split
the data into different groups and compute the summaries for each group.

```{r}
iris %>% 
    group_by(Species) %>%
    summarise(Mean.Petal.Length = mean(Petal.Length))
```

`n()` function counts how many observations you have in a subset of your data:

```{r}
iris %>%
    summarise(Observation = n())
```

Again this is more powerful combined with `group_by()`:
```{r}
iris %>%
    group_by(Species) %>%
    summarise(Number.Of.Species = n())
```

You can combine summary statistics simply by specifying more than one in the `summary()` function:
```{r}
iris %>%
    group_by(Species) %>%
    summarise(Number_of_Samples = n(),
              Mean.Petal.Length = mean(Petal.Length))
```






